Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2017-12-04 PAPER.tex
finished combing through entire paper for syntax/typos/clarity

\noindent Because of the nature of the lower triangular matrix, this window function has the property of preventing the leakage 
of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that 
power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low 
$k$'s, this method ensures a window function that retains clean, noise-dominated measurements while minimizing the 
contamination of foreground bias.

@@

The highest sensitivity power spectrum result for PAPER-64 using the updated analysis presented in this paper, shown in 
Figure \ref{fig:ps1_data}, has positive biases at low $k$ values. As discussed in Section \ref{sec:BiasTypes}, these detections 
are most likely attributable to foreground leakage. Here we demonstrate $3$ null tests performed on PAPER-64 data that verify 
that the positive detections are indeed due to foreground variation and not attributable to EoR.

The $3$ null test results are shown in Figure \ref{fig:null}, with each test described as the following:

\begin{itemize}
\item Original (black): This is identical to the power spectrum in Figure \ref{fig:ps1_data} and represents the best revised 
PAPER-64 power spectrum (one baseline type only) with weighting matrix $\hat{\textbf{C}}_{eff}$. There are clear detections at 
low $k$'s.
\item Even/Odd (blue): We split our dataset into even (e) and odd (o) days, where $\textbf{x}_{1} = e + o$ and $\textbf{x}_{2} = 
e - o$. We form datasets in this way to ensure that we use the full sensitivity of our data. When cross-multiplied, we obtain:
\begin{equation}
\textbf{x}_{1}\textbf{x}_{2}^{\dagger} = ee^{\dagger} - eo^{\dagger} + oe^{\dagger} - oo^{\dagger}
\end{equation}
If the same sky signal is in both the even and odd datasets, we expect it to cancel out.
\item Baselines (green): We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. Again, we form $\textbf{x}_{1} = b_{1} + b_{2}$ and $\textbf{x}_{2} = b_{1} - b_{2}$, where $b_{1}$ is the first baseline 
group and $b_{2}$ is the second baseline group.
\item LST (purple): We split our dataset into two halves based on LST, namely $t_{1}$ (LSTs 0.5-4.6 hours) and $t_{2}$ (LSTs 
4.6-8.6 hours). We form our two datasets as $\textbf{x}_{1} = t_{1} + t_{2}$ and $\textbf{x}_{2} = t_{1} - t_{2}$.
\end{itemize}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/null.png}
	\caption{Results for $3$ null tests compared to the PAPER-64 revised power spectrum (black) and analytically estimated 
$2\sigma$ errors (gray shaded region). For each, we took jackknives along different axes of the dataset - by julian days (separating 
even and odd days; blue), by baselines (green), and by LST (purple). We expect the sky signal to disappear for a `passing' null test. 
We find that the biases in our power spectrum are likely caused by variation of foregrounds in LST, and are not attributable to 
particular baselines or days of data.}
	\label{fig:null}
\end{figure}

@@

The null test in LST (purple) is much more illuminating. The fact that there are many power spectrum points outside the 
estimated error region signifies that the bias we see is likely from foregrounds that vary in LST. In other words, $t_{1}$ and 
$t_{2}$ differ by an amount greater than thermal noise and they do not effectively cancel out the sky signal during cross-
multiplication. The errors for this null test are also large, also reflective of residual foregrounds.

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances (Section 
\ref{sec:SiglossOverview}). Loss of the $21$ cm signal is especially significant if more eigenmodes are down-weighted than 
exist in the data (having an under-constrained covariance matrix). Hence, there exists a trade-off between sensitivity driven 
time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss (Section \ref{sec:Sigloss}). However, a 
signal-only simulation (i.e. comparing an unweighted vs. weighted power spectrum of EoR only) can underestimate loss by 
failing to account for spurious correlations between the data and signal.
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are significantly correlated 
(Section \ref{sec:ErrorOverview}). If the number of independent samples in a dataset is well-determined, bootstrapping is a 
simple and accurate way of estimating errors.
\item Meaningful null tests are vital to validate an EoR detection (Section \ref{sec:JackknifeOverview}). Similarly, performing 
jackknife tests along multiple axes of a dataset is necessary for confidence in an EoR detection and can also be used to tease 
out systematics.
\end{itemize}

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in P64, was underestimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim7$ mK higher than previous limits. Underestimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to underestimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were underestimated by a factor of $\sim$$2$ due to 
oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. They were also 
overestimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power 
spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the 
previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in P64 was too high for its data 
points.
\end{itemize}

The $21$ cm future is exciting, as new experiments have sensitivities that expect to reach and surpass EoR levels, improved 
foreground mitigation/removal strategies are being developed, and simulations are being designed to better understand 
instruments. On the power spectrum analysis side, robust signal loss simulations, precise error calculations, and 
comprehensive jackknife tests will play critical roles in accurate $21$ cm results. Looking ahead, the future promises to bring 
higher sensitivity measurements, better foreground and instrumental models, more effective and less lossy weighting 
techniques --- and with all that, an improved understanding of reionization and our early Universe.

@@

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). SAK is supported by the University of Pennsylvania School of Arts and Sciences Dissertation Completion Fellowship. We graciously thank SKA-
SA for site infrastructure and observing support.
\label{sec:Ack}

@@

The minimum-variance quadratic estimator $\hat{p}_\alpha$ for the $\alpha$th bandpower of the power spectrum is given by 
\begin{equation}
\hat{p}_\alpha = \frac{1} {2 \F_{\alpha \alpha} }\x^t \C^{-1} \Q^{\alpha} \C^{-1} \x,
\end{equation}
where
\begin{equation}
F_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \right)
\end{equation}
is the $\alpha$th diagonal element of the Fisher matrix\footnote{For this section only, with no loss of generality, we assume that the data $\textbf{x}$ are real.}. In our case, however, we do not have \emph{a priori} knowledge of the covariance matrix. Thus, we replace $\C$ with $\Chat$, its data-derived approximation. Our estimator then becomes
\begin{equation}
\label{eq:phatloss}
\hat{p}_\alpha^\textrm{loss} = \frac{1} {2 \hat{\F}_{\alpha \alpha} }\x^t \Chat^{-1} \Q^{\alpha} \Chat^{-1} \x,
\end{equation}
where
\begin{equation}
\hat{F}_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \Chat^{-1} \Q^\alpha \Chat^{-1} \Q^\alpha \right),
\end{equation}
with the label ``loss" to foreshadow the fact that this will be an estimator with signal loss (i.e., a multiplicative bias of less than unity). We will now provide an explicit demonstration of this by modeling the estimated covariance as
\begin{equation}
\label{eq:ChatDef}
\Chat = (1-\eta) \C + \eta \x \x^t,
\end{equation}
where $\eta$ is parameter quantifying our success at estimating the true covariance matrix. If $\eta = 0$, our covariance estimate has perfectly modeled the true covariance and $\Chat = \C$. On the other hand, if $\eta =1$, then our covariance estimate is based purely on the one realization of the covariance that is our actual data, and we would expect a high level of overfitting and signal loss.

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\hat{p}_\alpha^\textrm{loss}$ in terms of $\hat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\hat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\hat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading order in $\eta$. The result is
\begin{equation}
\hat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \hat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N) p_\alpha,
\end{equation}
where $N$ is the length of $\x$. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained. Now, recall that $p_\alpha$ is the \emph{true} power spectrum. This means that it can be decomposed into the sum of foreground and EoR power spectra, since the foregrounds and EoR are uncorrelated in expectation. This toy example, while not definitive, serves to motivate the form of Equation \eqref{eq:LinearPspecSum}.

\section{Simplified Loss Correction}
\label{sec:Pr_appendix}
One of the main difficulties in defining a signal loss procedure is in the interpretation of the injection procedure `output'. The method described in section \ref{sec:Sigloss} relies on several assumptions which required justification including the neglection of cross-terms in the power spectrum and indeed does not behave as expected at small injected values. However it is possible to make a simpler interpretation which requires less motivation.

In the main text we made an estimate of the signal loss effecting only the injected eor by supposing that the lossy injected signal was the difference between the injected and not injected cases: $ P(e) = P(x+e) - P(x)$. Suppose that we rephrase the question to: ``At what point would an injected signal be detectable at significant levels?". Instead of detangling the part of the output power spectrum dominated by the injected signal we can simply use $P(x+e)$, the weighted power spectrum of the data plus injected signal ($\hat{\textbf{p}}_{r}$ in Section \ref{sec:Sigloss}).

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{plots/method2_heatmap_I.png}
\caption{The distribution of output power spectrum values for a 
given injection level P$_{in}$. The $P_{\rm out}$ values that are plotted on the y-axis represent the weighted power spectrum of data plus the injected signal. It differs from that of Figure \ref{fig:sigloss_transfercurve} by not subtracting off the power spectrum of data alone (not subtracting off the second term in Equation \eqref{eq:sigloss}). We show kernel density estimations of the power spectrum transfer functions as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black 
diagonal line marks a perfect unity mapping, and the solid gray horizontal line denotes the peak of $\hat{\textbf{p}}_{x}$, the data distribution. At low injection, $P_{\rm out}$ converges at the data level.}
\label{fig:Pr_vs_Pin}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=.5\textwidth]{plots/method2_prob_vs_pinj.png}
\caption{The probability that an injected signal of level P$_{in}$
is observable at a 2$\sigma$ level above the value of $\hat{\textbf{p}}_{x}$ (data).
The different colors represent different $k$ values. We choose a probably of $90\%$ for our signal loss estimated power spectrum.
\label{fig:Prob_vs_Pin}}
\end{figure}

One difference in this method is that we are exclusively asking about signals detectable \emph{above} the level of the observed data. Thus it is explicitly limited to placing an upper limit, even if in some $k$-modes one has actual detections above the noise.  This might be seen as more conservative, however, a similar assumption has already been made when choosing to down-weight the data using itself. In such a down-weighting step, one has assumed that significant excess power is associated with a foreground residual. We note that a high SNR detection will require a different method.

Formally we can express our test as: ``What is the $P_{\rm in}$ such that
\begin{equation}
P_{r} > P_{x} \Lambda,
\end{equation}  
where $\Lambda$ represents a choice of threshold?'' Of course, both sides of this equation are actually a distribution of values associated with data variance and injected model variance. One way to interpret this test for the two distributions is to compute the probability that the two distributions are not\footnote{i.e. the null test} the same.

 \begin{equation}
\Prob(P_{in})  =  1 -  \int_{r=x} \Prob(P_x) \Prob(P_r,P_{in}) dr
 \end{equation}
 where $\Prob(P_x)$ is the probability of obtaining the power spectrum $\hat{\textbf{p}}_x$ found via bootstrapping (power spectrum of data only) and $\Prob(P_r,P_{in})$ is the probability of obtaining the summed power spectrum $\hat{\textbf{p}}_{r}$ given a range of input power levels, as sampled in Figure \ref{fig:Pr_vs_Pin}.

Probability vs. injection level is shown in Figure \ref{fig:Prob_vs_Pin}. At low injection levels the variation between $x$ and $r$ is such that the null test is ruled out to within $20$\%. This is the level expected from two distributions with $\sim$10 effective samples. 
%\dcj{checked with a little python test but needs a cite} 
As the injected signal approaches the level of the data, the probability of null rejection rises steeply.

@@

\begin{figure}[t]
\centering
\includegraphics[width=.4\textwidth]{plots/sigloss_method_comparison.png}
\caption{A comparison of the signal loss Section \ref{sec:Sigloss} method
against the method in this appendix at $90$\% confidence (red curve).
Also shown are the upper limits of the unweighted power spectrum
(blue dashed) and theoretical noise model (green). \label{fig:sigloss_compare} }
\end{figure}

>>>>

\noindent Because of the nature of the lower triangular matrix, this window function has the property of preventing the leakage 
of foreground power from low $k$ to high $k$ modes. Specifically, we order the elements in $\textbf{F}$ in such a way so that 
power can leak from high $k$ modes to low $k$ modes, but not vice versa. Since most foreground power shows up at low 
$k$'s, this method ensures a window function that retains clean, noise-dominated measurements while minimizing the 
contamination of foreground bias. For this paper, we compute $\textbf{M}$ as described above only for the inverse variance weighted PAPER-64 result (our revised result for one baseline type) shown in Figure \ref{fig:ps1_data}. For all other power spectrum estimates in this paper, we use $\textbf{M}=\textbf{I}$ for simplicity.

@@

The highest sensitivity power spectrum result for PAPER-64 using the updated analysis presented in this paper, shown in 
Figure \ref{fig:ps1_data}, has positive biases at low $k$ values. As discussed in Section \ref{sec:BiasTypes}, these detections 
are most likely attributable to foreground leakage. Here we demonstrate three null tests performed on PAPER-64 data that verify 
that the positive detections are indeed due to foreground variation and not attributable to EoR.

The three null test results are shown in Figure \ref{fig:null}, with each test described as the following:

\begin{itemize}
\item Original Data (black): This is identical to the power spectrum in Figure \ref{fig:ps1_data} and represents the best revised 
PAPER-64 power spectrum (one baseline type only) with weighting matrix $\hat{\textbf{C}}_{eff}$. There are clear detections at 
low $k$'s.
\item Even/Odd (blue): We split our dataset into even (e) and odd (o) days, where $\textbf{x}_{1} = e + o$ and $\textbf{x}_{2} = 
e - o$. We form datasets in this way to ensure that we use the full sensitivity of our data. When cross-multiplied, we obtain:
\begin{equation}
\textbf{x}_{1}\textbf{x}_{2}^{\dagger} = ee^{\dagger} - eo^{\dagger} + oe^{\dagger} - oo^{\dagger}
\end{equation}
If the same sky signal is in both the even and odd datasets, we expect it to cancel out.
\item Baselines (green): We split our dataset into two halves, where each contains half of the total baselines used in the 
analysis. Again, we form $\textbf{x}_{1} = b_{1} + b_{2}$ and $\textbf{x}_{2} = b_{1} - b_{2}$, where $b_{1}$ is the first baseline 
group and $b_{2}$ is the second baseline group.
\item LST (magenta): We split our dataset into two halves based on LST, namely $t_{1}$ (LSTs 0.5-4.6 hours) and $t_{2}$ (LSTs 
4.6-8.6 hours). We form our two datasets as $\textbf{x}_{1} = t_{1} + t_{2}$ and $\textbf{x}_{2} = t_{1} - t_{2}$.
\end{itemize}

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{plots/null.png}
	\caption{Results for three null tests compared to the PAPER-64 revised power spectrum (black) and analytically estimated 
$2\sigma$ errors (gray shaded region). For each, we take a jackknife along a different axis of the dataset - along julian days (separating 
even and odd days; blue), along baselines (green), and along LST (magenta). We expect the sky signal to disappear for a `passing' null test. 
We find that the biases in our power spectrum are likely caused by variation of foregrounds in LST, and are not attributable to 
particular baselines or days of data.}
	\label{fig:null}
\end{figure}

@@

The null test in LST (magenta) is much more illuminating. The fact that there are many power spectrum points outside the 
estimated error region signifies that the bias we see is likely from foregrounds that vary in LST. In other words, $t_{1}$ and 
$t_{2}$ differ by an amount greater than thermal noise and they do not effectively cancel out the sky signal during cross-multiplication. The errors for this null test are also large, also reflective of residual foregrounds.

@@

\begin{itemize}
\item Substantial signal loss can result when weighting data using empirically estimated covariances (Section 
\ref{sec:SiglossOverview}). Loss of the $21$ cm signal is especially significant if fewer modes
exist in the data. Hence, there exists a trade-off between sensitivity driven 
time-averaging techniques such as fringe-rate filtering and signal loss. 
\item Signal injection and recovery simulations can be used to quantify signal loss (Section \ref{sec:Sigloss}). However, a 
signal-only simulation (i.e. comparing an unweighted vs. weighted power spectrum of EoR only) can underestimate loss by 
failing to account for spurious correlations between the data and signal.
\item Errors that are estimated via bootstrapping can be underestimated if samples in the dataset are significantly correlated 
(Section \ref{sec:ErrorOverview}). If the number of independent samples in a dataset is well-determined, bootstrapping is a 
simple and accurate way of estimating errors.
\item Meaningful null tests are vital to validate an EoR detection (Section \ref{sec:JackknifeOverview}). Similarly, performing 
jackknife tests along multiple axes of a dataset is necessary for confidence in an EoR detection and can also be used to tease 
out systematics.
\end{itemize}

@@

\begin{itemize}
\item Signal loss, previously found to be $<2\%$ in P64, was under-estimated by a factor of $\sim$$1000$ for inverse 
covariance weighting. For our new analysis, we use a regularized covariance weighting method to minimize loss (Section 
\ref{sec:Weight}). However, because our revised weighting method is not as aggressive as the former, our results are still a 
factor of $\sim7$ mK higher than previous limits. Under-estimated signal loss therefore represents the bulk of our revision. We 
note that PAPER's analysis is not the first to under-estimate loss; results from the GMRT (\citealt{paciga_et_al2013}) were also 
revised from new signal loss calculations associated with their singular value decomposition foreground filter.
\item Power spectrum errors, originally computed by bootstrapping, were under-estimated by a factor of $\sim$$2$ due to 
oversampling data whose effective number of independent samples was reduced from fringe-rate filtering. Our limits were also 
over-estimated by a factor of $\sim$$2$ from using a suboptimal way of combining baselines into groups when forming power 
spectra (Section \ref{sec:Boot}). Hence, the net change for our error estimation is negligible, though the rationale behind the 
previous estimation was incorrect.
\item Several factors used in an analytic expression to predict the noise-level in PAPER-64 data were revised, yielding a 
decrease in predicted sensitivity level by a factor of $\sim$$2$ in mK (Section \ref{sec:PSSense}). We have verified our revised 
prediction extensively using pure noise simulations. We note that our sensitivity prediction is revised by a factor less than our 
power spectrum result, implying that if taken at face value, the theoretical prediction for noise in P64 was too high for its data 
points.
\end{itemize}

The $21$ cm future is exciting, as new experiments have sensitivities that expect to reach and surpass EoR levels, improved 
foreground mitigation/removal strategies are being developed, and simulations are being designed to better understand 
instruments. On the power spectrum analysis side, robust signal loss simulations, precise error calculations, and 
comprehensive jackknife tests will play critical roles in accurate $21$ cm results. With strong foundations being established now, it is safe to say that we can expect to learn much about reionization and our early Universe in the coming years.

@@

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship, National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400), and thank Eric Switzer for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). SAK is supported by the University of Pennsylvania School of Arts and Sciences Dissertation Completion Fellowship. We graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

@@

The minimum-variance quadratic estimator $\hat{p}_\alpha$ for the $\alpha$th bandpower of the power spectrum is given by 
\begin{equation}
\hat{p}_\alpha = \frac{1} {2 \F_{\alpha \alpha} }\x^t \C^{-1} \Q^{\alpha} \C^{-1} \x,
\end{equation}
where
\begin{equation}
F_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \right)
\end{equation}
is the $\alpha$th diagonal element of the Fisher matrix\footnote{For this section only, with no loss of generality, we assume that the data $\textbf{x}$ are real.}. In our case, however, we do not have \emph{a priori} knowledge of the covariance matrix. Thus, we replace $\C$ with $\Chat$, its data-derived approximation. Our estimator then becomes
\begin{equation}
\label{eq:phatloss}
\hat{p}_\alpha^\textrm{loss} = \frac{1} {2 \hat{\F}_{\alpha \alpha} }\x^t \Chat^{-1} \Q^{\alpha} \Chat^{-1} \x,
\end{equation}
where
\begin{equation}
\hat{F}_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \Chat^{-1} \Q^\alpha \Chat^{-1} \Q^\alpha \right),
\end{equation}
with the label ``loss" to foreshadow the fact that this will be an estimator with signal loss (i.e., a multiplicative bias of less than unity). We will now provide an explicit demonstration of this by modeling the estimated covariance as
\begin{equation}
\label{eq:ChatDef}
\Chat = (1-\eta) \C + \eta \x \x^t,
\end{equation}
where $\eta$ is a parameter quantifying our success at estimating the true covariance matrix. If $\eta = 0$, our covariance estimate has perfectly modeled the true covariance and $\Chat = \C$. On the other hand, if $\eta =1$, then our covariance estimate is based purely on the one realization of the covariance that is our actual data, and we would expect a high level of overfitting and signal loss.

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\hat{p}_\alpha^\textrm{loss}$ in terms of $\hat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\hat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\hat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading order in $\eta$. The result is
\begin{equation}
\hat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \hat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N) p_\alpha,
\end{equation}
where $N$ is the length of $\x$. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained. Now, recall that $p_\alpha$ is the \emph{true} power spectrum. This means that it can be decomposed into the sum of the foreground and EoR power spectra, since the foregrounds and EoR are uncorrelated in expectation. This toy example, while not definitive, serves to motivate the form of Equation \eqref{eq:LinearPspecSum}.

\section{Simplified Loss Correction}
\label{sec:Pr_appendix}
One of the main difficulties in defining a signal loss procedure is in the interpretation of the injection procedure `output'. The method described in Section \ref{sec:Sigloss} relies on several assumptions which required justification including the neglection of cross-terms in the power spectrum and indeed does not behave as expected at small injected values. However it is possible to make a simpler interpretation which requires less motivation.

In the main text we made an estimate of the signal loss effecting only the injected EoR by supposing that the lossy injected signal was the difference between the injected and not injected cases: $P_{out} = \hat{\textbf{p}}_{r} - \hat{\textbf{p}}_{x}$. Suppose that we rephrase the question we are asking to: ``At what point would an injected signal be detectable at significant levels?". Instead of detangling the part of the output power spectrum dominated by the injected signal we can simply use $\hat{\textbf{p}}_{r}$, the weighted power spectrum of the data plus injected signal.

\begin{figure}[tp]
\centering
\includegraphics[width=\textwidth]{plots/method2_heatmap_I.png}
\caption{The distribution of output power spectrum values for a 
given injection level P$_{in}$. The $P_{\rm out}$ values that are plotted on the y-axis represent the weighted power spectrum of data plus the injected signal. It differs from that of Figure \ref{fig:sigloss_transfercurve} by not subtracting off the power spectrum of data alone (not subtracting off the second term in Equation \eqref{eq:sigloss}). We show kernel density estimations of the power spectrum transfer functions as colored heat-maps for the cases of inverse covariance weighted PAPER-64 data (left) and unweighted data (right). The solid black 
diagonal line marks a perfect unity mapping, and the solid gray horizontal line denotes the peak of $\hat{\textbf{p}}_{x}$, the data distribution. At low injection, $P_{\rm out}$ converges to the data level.}
\label{fig:Pr_vs_Pin}
\end{figure}

@@

\begin{figure}[tp]
\centering
\includegraphics[width=.5\textwidth]{plots/method2_prob_vs_pinj.png}
\caption{The probability that an injected signal of level P$_{inj}$
is observable at a 2$\sigma$ level above the value of $\hat{\textbf{p}}_{x}$ (data).
The different colors represent different $k$ values. We choose a probably of $90\%$ for our signal loss estimated power spectrum.
\label{fig:Prob_vs_Pin}}
\end{figure}

One difference in this method from the one in the main text is that we are exclusively asking about signals detectable \emph{above} the level of the observed data. Thus it is explicitly limited to placing an upper limit, even if in some $k$-modes one has actual detections above the noise.  This might be seen as more conservative, however, a similar assumption has already been made when choosing to down-weight the data using itself. In such a down-weighting step, one has assumed that significant excess power is associated with a foreground residual. We note that a high SNR detection will require a different method.

Formally we can express our test as: ``What is the $P_{\rm in}$ such that
\begin{equation}
\hat{\textbf{p}}_{r} > \hat{\textbf{p}}_{x} \Lambda,
\end{equation}  
where $\Lambda$ represents a choice of threshold?'' Of course, both sides of this equation are actually a distribution of values associated with data variance and injected model variance. One way to interpret this test for the two distributions is to compute the probability that the two distributions are not\footnote{i.e. the null test} the same.

 \begin{equation}
\Prob(P_{in})  =  1 -  \int_{r=x} \Prob(\hat{\textbf{p}}_x) \Prob(\hat{\textbf{p}}_r,P_{in}) dr
 \end{equation}
 where $\Prob(\hat{\textbf{p}}_x)$ is the probability of obtaining the power spectrum $\hat{\textbf{p}}_x$ found via bootstrapping (power spectrum of data only) and $\Prob(\hat{\textbf{p}}_r,P_{in})$ is the probability of obtaining the summed power spectrum $\hat{\textbf{p}}_{r}$ given a range of input power levels, as sampled in Figure \ref{fig:Pr_vs_Pin}.

Probability vs. injection level is shown in Figure \ref{fig:Prob_vs_Pin}. At low injection levels the variation between $\textbf{x}$ and $\textbf{r}$ is such that the null test is ruled out to within $20$\%. This is the level expected from two distributions with $\sim$10 effective samples. 
%\dcj{checked with a little python test but needs a cite} 
As the injected signal approaches the level of the data, the probability of null rejection rises steeply.

@@

\begin{figure}[t]
\centering
\includegraphics[width=.4\textwidth]{plots/sigloss_method_comparison.png}
\caption{A comparison of the signal loss method described in Section \ref{sec:Sigloss}
with the method in this appendix at $90$\% confidence (red curve).
Also shown are the upper limits of the unweighted power spectrum
(blue dashed) and theoretical noise model (green). \label{fig:sigloss_compare} }
\end{figure}
